---
title: ''
author: ''
date: ''
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
setwd("C:/Users/orlyo/OneDrive/Desktop/Grad School/Spring 2021/2. STAT 2132 - Applied Stat. Methods 2/Homeworks/HW5")
c = read.csv("Q1c.csv")
c$Y = as.factor(c$Y)
d = read.csv("Q1d.csv")
d$Y = as.factor(d$Y)
library(MASS)
library(effects)
library(VGAM)
```

# (c)

*Using R, make two plots:*

*(i) On the same axes, plot the estimated probabilities from (Mi) and the estimated probabilities from (Mii) as a function of xi.*

```{r, fig.width = 6, fig.height = 4}
## Dataset c
fiti.c = glm(Yi ~ X, data = c, family = binomial(link = "logit"))
# summary(fiti.c)
fitii.c = glm(Yii ~ X, data = c, family = binomial(link = "logit"))
# summary(fitii.c)

# estimated probabilities
plot(c$X, fiti.c$fitted.values, col = "blue", pch = 20, main = "Estimated Probabilities - Dataset C", xlab = "X", ylab = "Fitted Values")
points(c$X, fitii.c$fitted.values, col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)
```

*(ii) On the same axes, plot the logit of the estimated probabilities from (Mi) and the logit of the estimated probabilities from (Mii) as a function of xi.*

```{r, fig.width = 6, fig.height = 4}
# logit of estimated probabilities
plot(c$X, log(fiti.c$fitted.values/(1 - fiti.c$fitted.values)), col = "blue", pch = 20, main = "Logit of Estimated Probabilities - Dataset C", xlab = "X", ylab = "Fitted Values")
points(c$X, log((fitii.c$fitted.values)/(1 - (fitii.c$fitted.values))), col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)
```

# (d)

*The file Q1d.txt on the Courseweb also contains a sample data set with n = 200. Repeat part (c) using this data set.*

```{r, fig.width = 6, fig.height = 4}
## Dataset d
fiti.d = glm(Yi ~ X, data = d, family = binomial(link = "logit"))
# summary(fiti.d)
fitii.d = glm(Yii ~ X, data = d, family = binomial(link = "logit"))
# summary(fitii.d)

# estimated probabilities
plot(d$X, fiti.d$fitted.values, col = "blue", pch = 20, main = "Estimated Probabilities - Dataset D", xlab = "X", ylab = "Fitted Values")
points(d$X, fitii.d$fitted.values, col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)

# logit of estimated probabilities
plot(d$X, log(fiti.d$fitted.values/(1 - fiti.d$fitted.values)), col = "blue", pch = 20, main = "Logit of Estimated Probabilities - Dataset D", xlab = "X", ylab = "Fitted Values")
points(d$X, log((fitii.d$fitted.values)/(1 - (fitii.d$fitted.values))), col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)
```








```{r, eval = FALSE}
m.c$probs = predict(m.c, type = "probs")
m.d$probs = predict(m.d, type = "probs")

# Probabilities
plot(effect("X", m.c))
plot(effect("X", m.c), style = "stacked")
plot(effect("X", m.c, latent = TRUE))

#############################################################################

fit = vglm(Y ~ X, family = cumulative(parallel = TRUE), data = c)

life1 = seq(0, 9, 1)
fit.logit.ses0 = predict(fit, newdata = data.frame(life = life1, ses = 0))
fit.logit.ses1 = predict(fit, newdata = data.frame(life = life1, ses = 1))
name = colnames(fit.logit.ses1)
plot.data = data.frame(life = rep(life1, 3), ses = rep(c("SES = low", "SES = high"), each = 3*10),
type = rep(name, each = 10), logit = c(fit.logit.ses0, fit.logit.ses1))
xyplot(Y ~ X, data = c, group = type, type = "l", auto.key = list(points = FALSE, lines = FALSE, columns = 3))

##############################################################################

library(sjPlot)
plot_model(m.c, type = "pred", show.ci = TRUE, terms = "X")
plot_model(fit, type = "pred", show.ci = TRUE, terms = "X")
```



```{r, eval = FALSE}
## Dataset c
fiti.c = glm(Yi ~ X, data = c, family = binomial(link = "logit"))
fitii.c = glm(Yii ~ X, data = c, family = binomial(link = "logit"))

# estimated probabilities
plot(c$X, fiti.c$fitted.values, col = "blue", pch = 20, main = "Estimated Probabilities - Dataset C", xlab = "X", ylab = "Fitted Values")
points(c$X, fitii.c$fitted.values, col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)

# logit of estimated probabilities
plot(c$X, log(fiti.c$fitted.values/(1 - fiti.c$fitted.values)), col = "blue", pch = 20, main = "Logit of Estimated Probabilities - Dataset C", xlab = "X", ylab = "Fitted Values")
points(c$X, log((fitii.c$fitted.values)/(1 - (fitii.c$fitted.values))), col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)


## Dataset d
fiti.d = glm(Yi ~ X, data = d, family = binomial(link = "logit"))
fitii.d = glm(Yii ~ X, data = d, family = binomial(link = "logit"))
# estimated probabilities
plot(d$X, fiti.d$fitted.values, col = "blue", pch = 20, main = "Estimated Probabilities - Dataset D", xlab = "X", ylab = "Fitted Values")
points(d$X, fitii.d$fitted.values, col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)

# logit of estimated probabilities
plot(d$X, log(fiti.d$fitted.values/(1 - fiti.d$fitted.values)), col = "blue", pch = 20, main = "Logit of Estimated Probabilities - Dataset D", xlab = "X", ylab = "Fitted Values")
points(d$X, log((fitii.d$fitted.values)/(1 - (fitii.d$fitted.values))), col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)

```


```{r, eval = FALSE}
## Dataset C
m.c = polr(Y ~ X, data = c, Hess = TRUE)
# summary(m.c)
# m.c$fitted.values

# plot for estimated probabilities
plot(c$X, m.c$fitted.values[,1], col = "blue", ylim = c(0,1), pch = 20, main = "Estimated Probabilities", xlab = "X", ylab = "Fitted Values")
points(c$X, (1 - m.c$fitted.values[,3]), col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)

# plot for logit of estimated probabilities
plot(c$X, log(m.c$fitted.values[,1]/(1 - m.c$fitted.values[,1])), col = "blue", pch = 20, main = "Logit of Estimated Probabilities", xlab = "X", ylab = "Fitted Values")
points(c$X, log((1 - m.c$fitted.values[,3])/(1 - ((1 - m.c$fitted.values[,3])))), col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)


## Dataset D
m.d = polr(Y ~ X, data = d, Hess = TRUE)
# summary(m.d)
# m.d$fitted.values

# plot for estimated probabilities
plot(d$X, m.d$fitted.values[,1], col = "blue", ylim = c(0,1), pch = 20, main = "Estimated Probabilities", xlab = "X", ylab = "Fitted Values")
points(d$X, (1 - m.d$fitted.values[,3]), col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)

# plot for logit of estimated probabilities
plot(d$X, log(m.d$fitted.values[,1]/(1 - m.d$fitted.values[,1])), col = "blue", pch = 20, main = "Logit of Estimated Probabilities", xlab = "X", ylab = "Fitted Values")
points(d$X, log((1 - m.d$fitted.values[,3])/(1 - ((1 - m.d$fitted.values[,3])))), col = "green", pch = 20)
legend("topleft", legend = c("Mi", "Mii"), col = c("blue", "green"), lty = 1, lwd = 2)



```

